---
title: "Project 2: Bank marketing dataset "
author: "Group 17"
number-sections: true
format: 
  pdf:
    fontsize: 12pt
    linestretch: 1.5
    geometry: margin=1in
    embed-resources: true
editor_options:
  chunk_output_type: console
---

```{r}
#| echo: false
#| warning: false
#| message: false

# load the libraries we will use.
packages <-c("skimr",
             "corrplot",
             "rpart",
             "gt",
             "dplyr",
             "ggplot2",
             "gridExtra",
             "caret",
            "rpart.plot",
            "randomForest",
            "caret",
            "pROC",
            "ROCR",
            "MASS"
            )  

# Library packages
for(pkg in packages){
  library(pkg, character.only = TRUE)
}
```

\thispagestyle{empty}
\newpage

# Introduction {#sec-intro}

The goal of this project is to predict whether clients will subscribe to a term deposit by analyzing the marketing data of the bank. The research findings of this report can assist financial institutions in formulating more effective marketing strategies, thereby increasing clients conversion rates and reducing marketing costs.

The data used in this study was sourced from a Portuguese bank and covered the detailed information of 17 marketing campaigns conducted by the bank from May 2008 to November 2010. This included the relevant data of 41,118 clients. The dataset included the basic characteristics of the clients, macroeconomic indicators, and variables related to the marketing campaigns. The response variable was a binary variable indicating whether the clients would subscribe to a term deposit ("Yes" or "No"). To enhance the efficiency of the analysis, 10,000 records were randomly selected from the original data as samples for modeling and analysis.

# Method {#sec-me}

To achieve the research objectives, this study will employ various classification methods, such as KNN, LDA, Tree-based methods, SVM, etc., to model the data and compare the models. These methods are applicable to binary classification problems and can also utilize multiple evaluation metrics to compare the predictive capabilities of the models. Moreover, through the comparison among different models, it is beneficial to better comprehensively understand the relationship between the data features and the prediction results.

The research results will be compared and analyzed using appropriate evaluation metrics, thereby identifying the model with the best performance and further explaining its predictive capabilities. Ultimately, this will provide a reference basis for future practical applications.

# Exploratory Analysis {#sec-EA}

```{r}
#| echo: false
#| warning: false
#| message: false
#| include: false

## read the data and take a quick look at the data

bank <- read.csv("group_17.csv", header=TRUE)

summary(bank)

#1. Target people for subscription: Primarily young and middle-aged individuals (with an average age of 40), the majority of whom are new  clients (never having been contacted before, most of pdays are 999).
#2. Call situation: There is an obvious variation in call duration (ranging from 1 to 82 (4918/60)minutes), indicating that there are differences in client responses;also, some clients have been contacted excessively (up to 37 times in the campaign).
#3. External economic background: The interest rate fluctuations in this data are relatively obvious (ranging of euribor3m from 0.6% to 5.0%), and sometimes changes in the external economic environment may affect clients' willingness to deposit.

skim(bank)

#1. Data completeness: There are no missing values in these 10,000 pieces of data (complete rate 100%), so the data can be directly used without the need for data filling.
#2. Client diversity: It includes 12 types of occupations and 8 levels of education . The target variable y is a binary classification (yes or no).
#3. Increasing new clients campaign: Over 75% of the clients ( pdays from the 25th percentile to the 100th percentile, all is 999) have never been contacted before, indicating that this is a marketing campaign to increase new clients.
#4. Client situation:
#Main client group: Concentrated in the age range of 32 to 47.
#Call situation: Half of the calls are no longer than 3 minutes (180 seconds).
#Interest rate situation: At the time of data recording, the market was in a high-interest-rate period (the median interest rate was 4.86%).

```

Before making a call, the variable `duration` is usually unknown, and this variable is highly correlated with the target variable. Specifically, it can only be determined whether the client subscribes after the call ends, and longer call duration is usually more likely to result in a subscription. Therefore, adding this variable to the model is equivalent to introducing "future information", which is logically unreasonable and the model has no practical value. Based on these considerations, the `duration` variable will be deleted to ensure that the model remains feasible in practical scenarios.

On the other hand,the dataset is highly imbalanced, with approximately 89% of observations belonging to the "no" class and around 11% to the "yes" class. A random seed (1234) was set, and the dataset was divided into training and test sets using an 80:20 ratio. The class distribution remains consistent after the split, with 88.84% “no” and 11.16% “yes” in the training set, and 88.6% “no” and 11.4% “yes” in the test set. It indicates that the random split has preserved the original class distribution and ensures that both the training and test sets are representative of the overall dataset. This is particularly important in the presence of class imbalance, as it helps avoid biased model evaluation and improves the reliability of performance comparisons across different models.

```{r}
#| echo: false
#| warning: false
#| message: false
#| include: false
#Data preprocessing

bank$duration <- NULL #Removing the "duration" variable is to make the model practical: this variable is unknown before making the call, and it is highly correlated with our target variable (we can know whether the client subscribes after the call, and clients with longer call durations are more likely to subscribe). If it is included in our model, it is equivalent to using "future information", which would lead to logically unreasonable predictions and make the constructed model of no practical value.

names(bank)

#Change the variable to a 1 or 0 variable indicating whether the client has been contacted or not
bank$pdays_contact <- ifelse(bank$pdays == 999, 0, 1)
bank$pdays <- NULL

# to change the character variables to factors for better fitting
bank$job <- as.factor(bank$job)
bank$marital <- as.factor(bank$marital)
bank$education <- as.factor(bank$education)
bank$default <- as.factor(bank$default)
bank$housing <- as.factor(bank$housing)
bank$loan <- as.factor(bank$loan)
bank$contact <- as.factor(bank$contact)
bank$month <- as.factor(bank$month)
bank$day_of_week <- as.factor(bank$day_of_week)
bank$poutcome <- as.factor(bank$poutcome)
bank$pdays_contact <- as.factor(bank$pdays_contact)
bank$y <- as.factor(bank$y)

```

```{r}
#| echo: false
#| warning: false
#| message: false
#| include: false

# split the data
# set seed
set.seed(1234)

# use 8:2 to split the data to two part (train data and test data)

train_index <- sample(1:nrow(bank), 0.8 * nrow(bank))

train_data <- bank[train_index, ]
test_data  <- bank[-train_index, ]

# check the dimensions
dim(train_data)
dim(test_data)

# check proportion of y
prop.table(table(train_data$y))
prop.table(table(test_data$y))

```

```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-cap: Distribution of Target Variable
#| label: fig-scat1
#| fig-align: center
#| fig-width: 6
#| fig-height: 4

#table(bank$y)

prop <- prop.table(table(train_data$y)) # to calculate the proportion

bp <- barplot(table(train_data$y), ylim=c(0,8000))
text(bp, table(train_data$y), paste0(round(prop*100, 1), "%"), pos = 3)#to add the proportion to the bar chart for better show in the report 

```

@fig-scat1 illustrates the uneven distribution of the target variable y: the proportion of non-subscribed clients reaches 88.8%, while the proportion of successfully converted clients is only 11.2%. Such class imbalance may introduce bias during model training. Therefore, model performance should not be evaluated solely based on overall accuracy, since even if a model predicts all results as "no", the overall accuracy can still exceed 80%. On the other hand, more attention will be paid to the model's ability to identify these 11.2% key samples, so that the final prediction results have practical reference value.

```{r}
#| echo: false
#| warning: false
#| message: false

# prepare data for correlation analysis (continuous variable)

bank_num <- train_data[,c("age",
                    "campaign",
                    "previous",
                    "emp.var.rate",
                    "cons.price.idx",
                    "cons.conf.idx",
                    "euribor3m",
                    "nr.employed")]

#head(bank_num)


```

```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-cap: Correlation Matrix of Numerical Variables
#| label: fig-scat2
#| fig-align: center
#| fig-width: 5
#| fig-height: 4.5

corrplot(cor(bank_num),
         method = "number",#low correlation will be close to white number
         type = "upper", #show half of the plot is enough
         order = "hclust",#to better show the most correlated part
         tl.col = "black",
         number.cex = 0.7)

```

```{r}
#| echo: false
#| warning: false
#| message: false
#| include: false

# The three variables (emp.var.rate, euribor3m, nr.employed) are highly correlated, so we use pca to decide to delete which two variables

bank.pca <- princomp(bank_num, cor = TRUE)

summary(bank.pca)

bank.pca$loadings

```

```{r}
#| echo: false
#| label: tbl-pca
#| tbl-cap: Principal Component Loadings for First Two

pca_loadings <- as.data.frame(unclass(bank.pca$loadings)) #use unclass() to delete NA

# add variable name
pca_loadings$variable <- rownames(pca_loadings)

# create the gt table
pca_loadings %>%
  dplyr::select(variable, Comp.1, Comp.2) %>%
  gt() %>%
  fmt_number(
    columns = c(Comp.1, Comp.2),
    decimals = 2
  )
```

By combining the correlation matrix @fig-scat2 and Principal Component Analysis @tbl-pca, it can be observed that `euribor3m`, `nr.employed`, and `emp.var.rate` are highly positively correlated (with pairwise correlation coefficients all above 0.9). Multicollinearity is also reflected in the first principal component (Comp.1). To improve subsequent modeling performance, `nr.employed` and `emp.var.rate` are excluded, while `euribor3m` is retained. As the Euro Interbank Offered Rate, `euribor3m` better captures the macro-financial environment during bank marketing and provides stronger interpretability for the final subscription outcomes. Meanwhile, since the variable `default` only has two values, "no" and "unknown", and over 79% of the cases are "no" with no instances of "yes", the available information is extremely limited. Therefore, this variable is removed to reduce noise in the future model.

```{r}
#| echo: false
#| warning: false
#| message: false

# delete nr.employed, and emp.var.rate variable

train_data$nr.employed <- NULL
train_data$emp.var.rate <- NULL
test_data$nr.employed <- NULL
test_data$emp.var.rate <- NULL

#names(train_data)
#names(test_data)
```

```{r}
#| echo: false
#| warning: false
#| message: false
#| include: false

# to have a look at categorical variables
# x and y are all categorical variables, so i use proportion to better show the relationship between them, to see who is more likely to subscribe( set  position = "fill" to show the stacked percentage bar chart)

ggplot(train_data, aes(job, fill = y)) + geom_bar(position = "fill")
# can see clear difference

ggplot(train_data, aes(marital, fill = y)) + geom_bar(position = "fill")
# table(train_data$marital)
# difference is not very clear. Just 13 records are "unknown". So it's possible most of this type is "no".

ggplot(train_data, aes(education, fill = y)) + geom_bar(position = "fill")
# can see clear difference

ggplot(train_data, aes(default, fill = y)) + geom_bar(position = "fill")
# difference is not very clear
# there is no "yes" in the default variable,the information is not enough,so i use table() to decide whether i need to delete it

table(train_data$default)

# Since the variable default only has two values, "no" and "unknown", and over 79% of the cases are "no" with no instances of "yes", the available information is extremely limited. Therefore, we have decided to remove this variable.

train_data$default <- NULL
test_data$default <- NULL

ggplot(train_data, aes(housing, fill = y)) + geom_bar(position = "fill")
# difference is not very clear

ggplot(train_data, aes(loan, fill = y)) + geom_bar(position = "fill")
# difference is not very clear

ggplot(train_data, aes(contact, fill = y)) + geom_bar(position = "fill")
# cellular might be more useful

ggplot(train_data, aes(month, fill = y)) + geom_bar(position = "fill")
# can see clear difference

ggplot(train_data, aes(day_of_week, fill = y)) + geom_bar(position = "fill")
# difference is not very clear

ggplot(train_data, aes(poutcome, fill = y)) + geom_bar(position = "fill")
# can see clear difference

ggplot(train_data, aes(pdays_contact, fill = y)) + geom_bar(position = "fill")
# can see clear difference


```

```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-cap: Subscription Rate Across Key Categorical Variables
#| label: fig-scat3
#| fig-align: center
#| fig-width: 6
#| fig-height: 4.5

# choose job, education, poutcome,contact variables to show their relationship with y，use ggplot to do it

p1 <- ggplot(train_data, aes(poutcome, fill = y)) +
  geom_bar(position = "fill") +
  labs(title = "Previous Outcome", x = "poutcome", y = "Proportion") +
  theme_minimal() + #use a simple theme
  theme(axis.text.x = element_text(angle = 45, hjust = 1))#to better show the x label ,rotate it


p2 <- ggplot(train_data, aes(job, fill = y)) +
  geom_bar(position = "fill") +
  labs(title = "Job", x = "job", y = "Proportion") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))#to better show the x label ,rotate it


p3 <- ggplot(train_data, aes(education, fill = y)) +
  geom_bar(position = "fill") +
  labs(title = "Education", x = "education", y = "Proportion") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))#to better show the x label ,rotate it

p4 <- ggplot(train_data, aes(contact, fill = y)) +
  geom_bar(position = "fill") +
  labs(title = "Contact Type", x = "contact", y = "Proportion") +
  theme_minimal()

grid.arrange(p1, p2, p3, p4, ncol = 2)
# use grid.arrange to put four graphs in one plot

```

@fig-scat3 shows the relationships between several key categorical variables and the variable indicating whether the client subscribes (y). Due to the large number of categorical variables, only selected variables with stronger interpretability and potential relevance to the target variable are presented. From the graphs, it can be seen that the previous marketing outcome (`poutcome`) obviously affects whether the client subscribes this time. Clients who were successfully marketed before (success) have a higher probability of subscribing than those who were marketing failed (failure) or had no previous contact (nonexistent). Moreover, there are certain differences between different occupations (`job`) and educational levels (`education`), indicating that the social characteristics of the clients may have an impact on the subscription decision. At the same time, there are also certain differences in contact methods (`contact`), with clients contacted via mobile phone (cellular) having a slightly higher chance of subscribing than those contacted via telephone.

Overall, historical campaign outcomes, client social characteristics, and communication channels may all have an impact on whether the client subscribes this time. Therefore, these variables will have certain predictive value in the subsequent modeling process.

```{r}
#| echo: false
#| warning: false
#| message: false
#| include: false

# to see the selected variables
names(train_data)
names(test_data)
```

```{r}
#| include: false
#| echo: false
#| label: tbl-var
#| tbl-cap: Variables Used in the Analysis

# This table presents the variables that we ultimately used for the analysis.

# use the information in Table 2: Variable description of bank marketing dataset to create a dataframe for better showing in the report .But it's too much. So we decided to not show it in the report, instead we use "Feature Selection Summary" to explain this part.

var_table <- data.frame(
  Variable = c(
    "y",
    "age","job","marital","education","housing","loan",
    "contact","month","day_of_week",
    "campaign","pdays_contact","previous","poutcome",
    "cons.price.idx","cons.conf.idx","euribor3m"
  ),
  
  Type = c(
    "Categorical",
    "Numeric","Categorical","Categorical","Categorical","Categorical","Categorical",
    "Categorical","Categorical","Categorical",
    "Numeric","Categorical","Numeric","Categorical",
    "Numeric","Numeric","Numeric"
  ),
  
  Description = c(
    "Has the client subscribed a term deposit (yes/no)",
    
    "Age of the client at the contact date",
    "Type of job (e.g., admin, technician, services, etc.)",
    "Marital status (divorced, married, single)",
    "Education level (basic.4y, basic.6y, basic.9y, high.school, university.degree, etc.)",
    "Whether the client has a housing loan",
    "Whether the client has a personal loan",
    
    "Contact communication type (cellular or telephone)",
    "Last contact month of the year",
    "Last contact day of the week",
    
    "Number of contacts performed during this campaign for this client",
    "Indicator whether the client has been contacted previously",
    "Number of contacts performed before this campaign",
    "Outcome of the previous marketing campaign (failure, nonexistent, success)",
    
    "Consumer price index (monthly indicator)",
    "Consumer confidence index (monthly indicator)",
    "Euribor 3-month rate (daily indicator)"
  )
)

# use gt() to better show the selected variable
var_table %>%
  gt() %>%
  tab_header(
    title = "Variables Used in the Analysis",
    subtitle = "Selected variables from the Bank Marketing dataset"
  )  %>% # add title and subtitle
  cols_label(
    Variable = "Variable Name",
    Type = "Variable Type",
    Description = "Description"
  ) #change the column label
```

### Feature Selection Summary {#sec-FSS .unnumbered}

The final set of variables includes clients characteristics (e.g., `age`, `job`, `education`), campaign-related variables (e.g., `campaign`, `previous`, `poutcome`), and macroeconomic indicators (e.g., `euribor3m`, consumer indices). `duration` was excluded to avoid data leakage, as it is only observable after a call has ended. `default` was removed due to near-zero variance, with no instances of "yes" observed in the 
training data. Additionally, multiple collinearity issues are addressed by retaining representative variables. Overall, the selected features lay a solid foundation for the subsequent modeling work.

# Formal Analysis {#sec-FA}

## Linear Discriminant Analysis (LDA)

### Model Specification

To avoid data leakage, `duration` was excluded from the analysis. For the LDA model, only continuous numerical predictors were included: `age`, `campaign`, `previous`, `cons.price.idx`, `cons.conf.idx`, and `euribor3m`, with `y` as the response variable.

As shown in @tbl-lda-var, a comparison of the group variances and covariance structures suggested that the equal covariance assumption was not fully satisfied. For example, the covariances for pairs such as `age` with `euribor3m` and `age` with `previous` differ across the two classes. LDA was therefore retained as a baseline discriminant model, while recognising that this assumption may be imperfect for the current dataset.

```{r}
#| echo: false
#| warning: false
#| message: false

# Select continuous numeric variables for LDA
lda_vars <- c("age", "campaign", "previous",
              "cons.price.idx", "cons.conf.idx", "euribor3m", "y")

train_lda <- train_data[, lda_vars]
test_lda  <- test_data[, lda_vars]

# Ensure target variable y is a factor variable
train_lda$y <- as.factor(train_lda$y)
test_lda$y  <- as.factor(test_lda$y)
```

```{r}
#| label: tbl-lda-var
#| tbl-cap: "Variance Comparison of Continuous Variables by Class in the Training Set"
#| echo: false
#| warning: false
#| message: false

num_vars <- c("age", "campaign", "previous",
              "cons.price.idx", "cons.conf.idx", "euribor3m")

var_summary <- data.frame(
  Variable = num_vars,
  Var_no = sapply(train_lda[train_lda$y == "no", num_vars], var),
  Var_yes = sapply(train_lda[train_lda$y == "yes", num_vars], var)
)

var_summary %>%
  gt() %>%
  fmt_number(columns = c(Var_no, Var_yes), decimals = 3) %>%
  tab_options(
    table.font.size = px(12)
  )
```

```{r}
#| label: lda-cov-check
#| echo: false
#| warning: false
#| message: false
#| results: false

cov_no  <- cov(train_lda[train_lda$y == "no", num_vars])
cov_yes <- cov(train_lda[train_lda$y == "yes", num_vars])

round(cov_no, 3)
round(cov_yes, 3)
```

### Model fitting

After standardising the predictors using the training-set mean and standard deviation, the LDA model was fitted to the training data. The LD1 coefficients suggest that `euribor3m`, `cons.price.idx`, and `cons.conf.idx` contributed most to class separation, while `campaign` and `age` had relatively small contributions (@tbl-lda-coefs).

```{r}
#| echo: false
#| warning: false
#| message: false

train_lda_scaled <- train_lda
test_lda_scaled  <- test_lda

train_means <- sapply(train_lda_scaled[, num_vars], mean)
train_sds   <- sapply(train_lda_scaled[, num_vars], sd)

train_lda_scaled[, num_vars] <- scale(train_lda_scaled[, num_vars],
                                      center = train_means,
                                      scale = train_sds)

test_lda_scaled[, num_vars] <- scale(test_lda_scaled[, num_vars],
                                     center = train_means,
                                     scale = train_sds)
```

```{r}
#| label: tbl-lda-coefs
#| tbl-cap: "Linear Discriminant Coefficients (LD1)"
#| tbl-pos: "H"
#| echo: false
#| warning: false
#| message: false

lda_model <- lda(y ~ ., data = train_lda_scaled)

ld1_coefs <- data.frame(
  Variable = rownames(lda_model$scaling),
  LD1_Coefficient = lda_model$scaling[, 1]
)

ld1_coefs %>%
  arrange(desc(abs(LD1_Coefficient))) %>%
  gt() %>%
  fmt_number(columns = LD1_Coefficient, decimals = 4
  )
```

```{r}
#| echo: false
#| warning: false
#| message: false

# LDA prediction on test set
lda_pred_test  <- predict(lda_model, newdata = test_lda_scaled)
lda_class_test <- lda_pred_test$class

# Confusion matrix
cm_lda <- confusionMatrix(lda_class_test, test_lda_scaled$y, positive = "yes")

# AUC
lda_posterior_yes <- lda_pred_test$posterior[, "yes"]
roc_lda <- roc(test_lda_scaled$y, lda_posterior_yes, quiet = TRUE)

# F1
precision_lda <- cm_lda$byClass["Pos Pred Value"]
recall_lda    <- cm_lda$byClass["Sensitivity"]
f1_lda        <- 2 * precision_lda * recall_lda / (precision_lda + recall_lda)



lda_results <- data.frame(
  Model = "LDA",
  Accuracy = unname(cm_lda$overall["Accuracy"]),
  Balanced_Accuracy = unname(cm_lda$byClass["Balanced Accuracy"]),
  Sensitivity = unname(cm_lda$byClass["Sensitivity"]),
  Specificity = unname(cm_lda$byClass["Specificity"]),
  Precision = unname(cm_lda$byClass["Pos Pred Value"]),
  F1_Score = unname(f1_lda),
  AUC = as.numeric(roc_lda$auc)
)
```

## Quadratic Discriminant Analysis(QDA)

### Model Description and Justification

The QDA model uses the same set of continuous predictors as the LDA model: `age`, `campaign`, `previous`, `cons.price.idx`, `cons.conf.idx`, and `euribor3m`, with `y` as the response variable. The variable duration was excluded earlier because it would introduce data leakage, so it is not included here. To ensure comparability across models, the training and test sets use the same standardised variables as in the LDA analysis.

Unlike LDA, QDA does not assume a common covariance matrix across classes. Since the earlier variance and covariance checks suggested that the two classes differ in their covariance structure, QDA provides a more flexible alternative and is therefore a reasonable extension of the LDA model for this dataset.

```{r}
#| echo: false
#| warning: false
#| message: false
#| results: false

# Fit QDA model
qda_model <- qda(y ~ ., data = train_lda_scaled)

# View model results
qda_model
```

The fitted QDA model also reflects the class imbalance observed in the data. The prior probabilities for the "no" and "yes" classes are 0.8884 and 0.1116 respectively, showing that non-subscribers make up the large majority of the sample. In addition, the group means suggest that "previous" tends to be higher in the "yes" class, while `euribor3m` tends to be lower. This indicates that previous contact history and macroeconomic conditions remain important for distinguishing between the two outcome classes.

```{r}
#| echo: false
#| warning: false
#| message: false

# QDA prediction on test set
qda_pred_test  <- predict(qda_model, newdata = test_lda_scaled)
qda_class_test <- qda_pred_test$class

# Confusion matrix
cm_qda <- confusionMatrix(qda_class_test, test_lda_scaled$y, positive = "yes")

# AUC
qda_posterior_yes <- qda_pred_test$posterior[, "yes"]
roc_qda <- roc(test_lda_scaled$y, qda_posterior_yes, quiet = TRUE)

# F1
precision_qda <- cm_qda$byClass["Pos Pred Value"]
recall_qda    <- cm_qda$byClass["Sensitivity"]
f1_qda        <- 2 * precision_qda * recall_qda / (precision_qda + recall_qda)


qda_results <- data.frame(
  Model = "QDA",
  Accuracy = unname(cm_qda$overall["Accuracy"]),
  Balanced_Accuracy = unname(cm_qda$byClass["Balanced Accuracy"]),
  Sensitivity = unname(cm_qda$byClass["Sensitivity"]),
  Specificity = unname(cm_qda$byClass["Specificity"]),
  Precision = unname(cm_qda$byClass["Pos Pred Value"]),
  F1_Score = unname(f1_qda),
  AUC = as.numeric(roc_qda$auc)
)
```

## Decision Tree

### Model Description and Justification

A decision tree classifier was applied as the first tree-based method. Decision trees recursively partition the feature space by selecting the variable and split point that best separates the classes at each node, measured by the Gini impurity criterion. The resulting model is fully interpretable, as its predictions can be explained through a sequence of human-readable if-then rules. This is especially useful in a banking context, where understanding why a client is predicted to subscribe is of clear practical importance.

Given the substantial class imbalance in the dataset (88.8% "no"), equal class priors were assigned (`prior = c(0.5, 0.5)`) to prevent the tree from trivially favouring the majority class. Without this adjustment, a tree predicting "no" for every observation would achieve over 88% accuracy while providing no useful predictive information.

```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-cap: "CP Plot for Decision Tree Pruning"
#| label: fig-tree-cp
#| fig-align: center
#| fig-width: 6
#| fig-height: 4

set.seed(1234)

# Build the full tree with class weights to handle imbalance
tree_full <- rpart(y ~ ., data = train_data,
                   method = "class",
                   parms = list(prior = c(0.5, 0.5)),  # equal prior to handle imbalance
                   control = rpart.control(cp = 0))

# CP plot (shown in report)
plotcp(tree_full)
```

### Model fitting and pruning

A full unpruned tree was first grown with the complexity parameter `cp = 0`, allowing the tree to expand until no further splits were possible. Unpruned trees are prone to overfitting — they learn noise specific to the training data and generalise poorly to unseen observations. Pruning is therefore necessary to find the right balance between bias and variance. The optimal `cp` was selected using the 1-SE rule: rather than choosing the `cp` corresponding to the absolute minimum cross-validation error (which tends to favour overly complex trees), the largest `cp` whose cross-validation error falls within one standard error of the minimum was selected. As shown in @fig-tree-cp, the cross-validation error curve flattens quickly after approximately 35 nodes, indicating that additional complexity yields negligible improvement. The 1-SE rule selected a tree with only 3 leaf nodes, which is both parsimonious and less prone to overfitting.

```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-cap: "Pruned Decision Tree"
#| label: fig-tree-plot
#| fig-align: center
#| fig-width: 4.5
#| fig-height: 3

# Use 1-SE rule to pick the smallest tree within 1 SE of minimum xerror
cp_table <- tree_full$cptable
min_xerror <- min(cp_table[, "xerror"])
se_threshold <- min_xerror + cp_table[which.min(cp_table[, "xerror"]), "xstd"]
best_cp <- cp_table[cp_table[, "xerror"] <= se_threshold, "CP"][1]

tree_pruned <- prune(tree_full, cp = best_cp)

# Visualisation
rpart.plot(tree_pruned, 
           type = 4, 
           extra = 104,
           cex = 1,
           box.palette = c("#C6DBEF", "#74C476"),
           shadow.col = "gray80",
           border.col = "white",
           split.font = 1,
           split.cex = 1)

```

### Tree structure and interpretation

The pruned tree (@fig-tree-plot) identifies two variables as the most informative predictors: First split — `euribor3m >= 1.3`: The Euribor 3-month rate, a macro-economic indicator reflecting the prevailing interest rate environment, forms the root split. When `euribor3m < 1.3` (low interest rate environment), the tree predicts "yes" with 83% probability, covering 14% of the test data. This is economically intuitive: when market interest rates are low, term deposits offer relatively more attractive returns, increasing clients' willingness to subscribe. Second split — `month`: Among observations where `euribor3m >= 1.3`, the contact month further differentiates clients. Contacts made in April, March, and October yield a 68% subscription rate, compared to 31% in the remaining months. This may reflect seasonal patterns in clients' financial planning or campaign timing effects. The simplicity of the final tree with two splits and three rules is both a strength and a limitation. It is highly interpretable and robust to overfitting, but it necessarily ignores the predictive information contained in the remaining variables.

```{r}
#| echo: false
#| warning: false
#| message: false

# Prediction
pred_tree <- predict(tree_pruned, test_data, type = "class")
prob_tree  <- predict(tree_pruned, test_data, type = "prob")[, "yes"]

# Confusion matrix and ROC
cm_tree  <- confusionMatrix(pred_tree, test_data$y, positive = "yes")
roc_tree <- roc(test_data$y, prob_tree, quiet = TRUE)

# F1 score
precision_tree <- cm_tree$byClass["Pos Pred Value"]
recall_tree    <- cm_tree$byClass["Sensitivity"]
f1_tree        <- 2 * precision_tree * recall_tree / (precision_tree + recall_tree)

tree_results <- data.frame(
  Metric = c("Accuracy", "Balanced Accuracy", "Sensitivity",
             "Specificity", "Precision", "F1 Score", "AUC"),
  Value = c(
    round(cm_tree$overall["Accuracy"], 4),
    round(cm_tree$byClass["Balanced Accuracy"], 4),
    round(cm_tree$byClass["Sensitivity"], 4),
    round(cm_tree$byClass["Specificity"], 4),
    round(precision_tree, 4),
    round(f1_tree, 4),
    round(auc(roc_tree), 4)
  )
)

```

## Random Forest

```{r}
#| echo: false
#| warning: false
#| message: false
#| include: false
set.seed(1234)

# Default model to check OOB error
rf_default <- randomForest(y ~ ., 
                           data = train_data,
                           ntree = 500,
                           importance = TRUE)

```

### Model description and justification

Random Forest is an ensemble learning method that constructs a large number of decision trees, each trained on a bootstrap sample of the training data. At each split, only a randomly selected subset of features is considered, which decorrelates the trees and reduces the variance of the ensemble compared to a single decision tree. Predictions are made by majority vote across all trees. This combination of bagging and feature randomisation makes Random Forest particularly robust to overfitting and well suited to tabular classification problems with mixed variable types.

```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-cap: "OOB Error vs mtry for Random Forest Tuning"
#| label: fig-rf-tune
#| fig-align: center
#| fig-width: 6
#| fig-height: 4
#| results: hide
# Tune mtry
set.seed(1234)

capture.output(
  tuneRF(
    train_data[, -which(names(train_data) == "y")],
    train_data$y,
    stepFactor = 1.5,
    improve = 0.01,
    ntreeTry = 200
  )
)

```

### Hyperparameter tuning

The key hyperparameter in Random Forest is `mtry`, the number of variables randomly considered at each split. The default value in R for classification is $\lfloor\sqrt{p}\rfloor$, where $p$ is the number of predictors, giving `mtry = 4` in this case. To find the optimal value, `tuneRF` was applied, evaluating each candidate using Out-of-Bag (OOB) error on the training data. As shown in @fig-rf-tune, OOB error is minimised at `mtry = 2` (OOB error = 9.56%), with error increasing as `mtry` grows.

```{r}
#| echo: false
#| warning: false
#| message: false
#| include: false 

# Final model with best mtry and class weights
set.seed(1234)
rf_final <- randomForest(y ~ ., data = train_data,
                         ntree = 500,
                         mtry = 2,  
                    # mtry=2 has lowest OOB error
                         importance = TRUE)

prob_rf <- predict(rf_final, test_data, type = "prob")[, "yes"]
roc_rf  <- roc(test_data$y, prob_rf, quiet = TRUE)           
                          
# Find the best threshold to maximize sensitivity and specificity
best_thresh <- coords(roc_rf, "best", 
                      best.method = "youden",
                      ret = "threshold")$threshold

invisible(cat("Optimal threshold:", round(best_thresh, 4), "\n"))

```

### Handling class imbalance

The default classification threshold of 0.5 is poorly suited to imbalanced data, as it systematically under-predicts the minority class. Adjusting class weights was explored but found to reduce AUC by distorting the model's probability estimates. Instead, the threshold was optimised after fitting using Youden's J statistic, which identifies the value that maximises the sum of Sensitivity and Specificity simultaneously. The optimal threshold was 0.083, which reflects the low base rate of subscription (11.4%) in the dataset.

```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-cap: "Variable Importance Plot - Random Forest (Mean Decrease in Accuracy and Gini)"
#| label: fig-rf-varimp
#| fig-align: center
#| fig-width: 6
#| fig-height: 4


# Variable importance plot
importance_df <- as.data.frame(importance(rf_final))
importance_df$variable <- rownames(importance_df)

p_acc <- ggplot(importance_df,
                aes(x = MeanDecreaseAccuracy,
                    y = reorder(variable, MeanDecreaseAccuracy))) +
  geom_point(color = "steelblue", size = 2.5) +
  labs(title = "Mean Decrease Accuracy",
       x = "Mean Decrease Accuracy", y = NULL) +
  theme_minimal(base_size = 11)

p_gini <- ggplot(importance_df,
                 aes(x = MeanDecreaseGini,
                     y = reorder(variable, MeanDecreaseGini))) +
  geom_point(color = "darkorange", size = 2.5) +
  labs(title = "Mean Decrease Gini",
       x = "Mean Decrease Gini", y = NULL) +
  theme_minimal(base_size = 11)

grid.arrange(p_acc, p_gini, ncol = 2,
             top = "Variable Importance - Random Forest")

```

### Variable importance

@fig-rf-varimp presents variable importance under two measures. Mean Decrease in Accuracy captures how much predictive performance drops when a variable's values are randomly permuted, providing a direct measure of its contribution to model accuracy. Mean Decrease in Gini reflects the total reduction in node impurity attributable to each variable across all trees.

Both measures consistently rank `euribor3m` as the most important predictor, reinforcing the finding from the decision tree that the prevailing interest rate environment is the strongest driver of subscription behaviour. `month` and `age` rank highly under Mean Decrease in Accuracy, while `age` and `job` are prominent under Mean Decrease in Gini. The two measures do not always agree because they capture different aspects of importance: permutation accuracy is more reliable when predictors are correlated, while Gini importance can favour variables with many distinct values. Taken together, the results suggest that macro-economic conditions, client demographics, and campaign timing are the most informative predictors. Variables such as `loan`, `housing`, and `contact` rank near the bottom on both measures, indicating limited independent predictive value once stronger predictors are accounted for.

```{r}
#| echo: false
#| warning: false
#| message: false

# Prediction
pred_rf_opt <- factor(ifelse(prob_rf >= best_thresh, "yes", "no"),
                      levels = c("no", "yes"))

# Confusion matrix and ROC
cm_rf <- confusionMatrix(pred_rf_opt, test_data$y, positive = "yes")

# F1 score
precision_rf <- cm_rf$byClass["Pos Pred Value"]
recall_rf    <- cm_rf$byClass["Sensitivity"]
f1_rf        <- 2 * precision_rf * recall_rf / (precision_rf + recall_rf)

rf_results <- data.frame(
  Metric = c("Accuracy", "Balanced Accuracy", "Sensitivity",
             "Specificity", "Precision", "F1 Score", "AUC"),
  Value = c(
    round(cm_rf$overall["Accuracy"], 4),
    round(cm_rf$byClass["Balanced Accuracy"], 4),
    round(cm_rf$byClass["Sensitivity"], 4),
    round(cm_rf$byClass["Specificity"], 4),
    round(precision_rf, 4),
    round(f1_rf, 4),
    round(auc(roc_rf), 4)
  )
)
```

## k-Nearest Neighbours (KNN)

```{r}
#| echo: false
#| warning: false
#| message: false

# Shared preprocessing for KNN and SVM
# These two methods require numerical inputs, so factor variables are converted
# into dummy variables and then standardised using training-set statistics only.

x_train_raw <- subset(train_data, select = -y)
x_test_raw  <- subset(test_data,  select = -y)

# create dummy variables from factor predictors
dummy_obj <- dummyVars(~ ., data = x_train_raw)

x_train_mat <- predict(dummy_obj, newdata = x_train_raw)
x_test_mat  <- predict(dummy_obj, newdata = x_test_raw)

# standardise using training set mean and sd
preproc_obj <- preProcess(x_train_mat, method = c("center", "scale"))

x_train_scaled <- predict(preproc_obj, x_train_mat)
x_test_scaled  <- predict(preproc_obj, x_test_mat)

# outcome variables
y_train <- train_data$y
y_test  <- test_data$y

```

### Model description and justification

The k-Nearest Neighbours (KNN) classifier is a non-parametric method that assigns class labels based on the majority class among the k closest observations in the predictor space. As KNN relies on distance calculations, all predictors were converted into numerical form through dummy encoding and subsequently standardised to ensure comparability across variables.

KNN was included in this study as a flexible, data-driven benchmark that does not impose a parametric structure on the relationship between predictors and the response. This makes it particularly useful for capturing local patterns in the data and for comparison with more structured classification methods.

```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-cap: "Cross-Validated ROC for KNN Across Candidate k Values"
#| label: fig-knn-tune
#| fig-align: center
#| fig-width: 6
#| fig-height: 4
set.seed(1234)

y_train_knn <- factor(y_train, levels = c("yes", "no"))

ctrl_knn <- trainControl(
  method = "cv",
  number = 5,
  classProbs = TRUE,
  summaryFunction = twoClassSummary
)

knn_grid <- expand.grid(k = c(3, 5, 7, 9, 11, 15, 21))

knn_fit <- train(
  x = x_train_scaled,
  y = y_train_knn,
  method = "knn",
  metric = "ROC",
  tuneGrid = knn_grid,
  trControl = ctrl_knn
)

best_k   <- knn_fit$bestTune$k
best_roc <- knn_fit$results$ROC[knn_fit$results$k == best_k]

ggplot(knn_fit$results, aes(x = k, y = ROC)) +
  geom_line(linewidth = 0.8, colour = "black") +
  geom_point(size = 2, colour = "black") +
  geom_point(
    data = subset(knn_fit$results, k == best_k),
    colour = "#d95f02",
    size = 3
  ) +
  geom_text(
    data = subset(knn_fit$results, k == best_k),
    aes(label = paste0("Best k = ", best_k)),
    vjust = 2,
    hjust = 0.9,
    colour = "#d95f02",
    size = 4
  ) +
  theme_minimal(base_size = 12) +
  labs(
    title = "KNN Tuning",
    x = "Number of Neighbours (k)",
    y = "Cross-Validated ROC"
  ) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    panel.grid.minor = element_blank()
  )
```

### Hyperparameter tuning

The primary tuning parameter for KNN is the number of neighbours, k. To determine an appropriate value, a range of candidate values was evaluated using cross-validated ROC as the selection criterion. This approach ensures that the chosen model maximises its ability to discriminate between subscribed and non-subscribed clients.

The results indicate that model performance and stability improve as k increases. The highest cross-validated ROC was achieved at k = 21, which was therefore selected as the optimal hyperparameter. Although performance differences beyond k = 9 were relatively small, selecting k = 21 ensures that the model achieves its best observed predictive performance.

```{r}
#| echo: false
#| warning: false
#| message: false

# Predictions on test set
pred_knn <- predict(knn_fit, x_test_scaled)
prob_knn <- predict(knn_fit, x_test_scaled, type = "prob")[, "yes"]

pred_knn <- factor(pred_knn, levels = c("no", "yes"))

cm_knn  <- confusionMatrix(pred_knn, y_test, positive = "yes")
roc_knn <- roc(y_test, prob_knn, levels = c("no", "yes"), quiet = TRUE)

# F1 score
precision_knn <- cm_knn$byClass["Pos Pred Value"]
recall_knn    <- cm_knn$byClass["Sensitivity"]
f1_knn        <- 2 * precision_knn * recall_knn / (precision_knn + recall_knn)

knn_results <- data.frame(
  Metric = c("AUC", "Sensitivity", "Specificity",
             "Balanced Accuracy", "F1 Score"),
  Value = c(
    round(auc(roc_knn), 4),
    round(cm_knn$byClass["Sensitivity"], 4),
    round(cm_knn$byClass["Specificity"], 4),
    round(cm_knn$byClass["Balanced Accuracy"], 4),
    round(f1_knn, 4)
  )
)

```

## Support Vector Machines (SVM)

### Model description and justification

Support Vector Machines (SVM) classify observations by identifying an optimal separating boundary that maximises the margin between classes. In this study, a radial basis function (RBF) kernel was employed to allow for nonlinear decision boundaries, enabling the model to capture complex relationships between predictors and the response variable.

SVM was selected due to its strong performance in high-dimensional settings and its ability to construct stable classification boundaries after dummy encoding of categorical variables. Compared with KNN, SVM provides a more structured approach to classification and can offer improved generalisation in the presence of noisy predictors.

```{r}
#| echo: false
#| warning: false
#| message: false
#| results: hide

set.seed(1234)

y_train_svm <- factor(y_train, levels = c("yes", "no"))

ctrl_svm <- trainControl(
  method = "cv",
  number = 3,
  classProbs = TRUE,
  summaryFunction = twoClassSummary
)

svm_grid <- expand.grid(
  sigma = c(0.01, 0.05),
  C     = c(0.5, 1, 2)
)

svm_fit <- train(
  x = x_train_scaled,
  y = y_train_svm,
  method = "svmRadial",
  metric = "ROC",
  tuneGrid = svm_grid,
  trControl = ctrl_svm
)


capture.output({
  svm_fit$bestTune
  svm_fit$results
})

```

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-SVM-turning
#| fig-cap: "SVM Tuning"
#| fig-align: center
#| fig-width: 5
#| fig-height: 3

ggplot(svm_fit$results, aes(x = factor(C), y = factor(sigma), fill = ROC)) +
  geom_tile(color = "white") +
  theme_minimal() +
  labs(x = "Cost (C)",
       y = "Sigma",
       fill = "ROC")
```

@fig-SVM-turning illustrates the cross-validated ROC values for different combinations of the SVM hyperparameters, namely the cost parameter (C) and the kernel parameter (sigma). Each cell represents the predictive performance of a specific parameter combination, with darker shading indicating higher ROC values.

It can be observed that models with a larger sigma (0.05) consistently achieve higher ROC values compared to those with sigma = 0.01, suggesting that a more flexible decision boundary improves classification performance in this dataset. Among all combinations, the highest ROC is obtained at sigma = 0.05 and C = 0.5, indicating that this parameter setting provides the best trade-off between model complexity and generalisation ability.

### Hyperparameter tuning

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-svm-best
#| tbl-cap: "SVM: Selected Hyperparameters"

svm_best <- data.frame(
  Hyperparameter = c("Best sigma", "Best cost (C)"),
  Value = c(svm_fit$bestTune$sigma, svm_fit$bestTune$C)
)

svm_best %>%
  gt()
```

The performance of the tuned SVM model was assessed on the test set using the same evaluation metrics as for KNN. The model achieved high overall accuracy, but, similar to KNN, this result is largely driven by the dominance of the majority class.

The relatively low sensitivity indicates that the SVM model struggles to correctly identify subscribed clients, suggesting that the decision boundary may favour the majority class. The AUC value provides a more balanced assessment of performance and indicates that the SVM model achieves moderate discrimination ability, though slightly lower than that of KNN in this application.

```{r}
#| echo: false
#| warning: false
#| message: false

pred_svm <- predict(svm_fit, x_test_scaled)
prob_svm <- predict(svm_fit, x_test_scaled, type = "prob")[, "yes"]

pred_svm <- factor(pred_svm, levels = c("no", "yes"))

cm_svm  <- confusionMatrix(pred_svm, y_test, positive = "yes")
roc_svm <- roc(y_test, prob_svm, levels = c("no", "yes"), quiet = TRUE)

# F1 score
precision_svm <- cm_svm$byClass["Pos Pred Value"]
recall_svm    <- cm_svm$byClass["Sensitivity"]
f1_svm        <- 2 * precision_svm * recall_svm / (precision_svm + recall_svm)

svm_results <- data.frame(
  Metric = c("AUC", "Sensitivity", "Specificity",
             "Balanced Accuracy", "F1 Score"),
  Value = c(
    round(auc(roc_svm), 4),
    round(cm_svm$byClass["Sensitivity"], 4),
    round(cm_svm$byClass["Specificity"], 4),
    round(cm_svm$byClass["Balanced Accuracy"], 4),
    round(f1_svm, 4)
  )
)

```

# Model Comparision and Selection {#sec-comp}

## Summary Table

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-comparision
#| tbl-cap: "Model Comparison on Test Set"
library(pROC)
library(gt)

results <- data.frame(
  Model = c("LDA", "QDA", "Decision Tree", "Random Forest", "KNN", "SVM"),
  
  Accuracy = c(
    as.numeric(cm_lda$overall["Accuracy"]),
    as.numeric(cm_qda$overall["Accuracy"]),
    as.numeric(cm_tree$overall["Accuracy"]),
    as.numeric(cm_rf$overall["Accuracy"]),
    as.numeric(cm_knn$overall["Accuracy"]),
    as.numeric(cm_svm$overall["Accuracy"])
  ),
  
  Sensitivity = c(
    as.numeric(cm_lda$byClass["Sensitivity"]),
    as.numeric(cm_qda$byClass["Sensitivity"]),
    as.numeric(cm_tree$byClass["Sensitivity"]),
    as.numeric(cm_rf$byClass["Sensitivity"]),
    as.numeric(cm_knn$byClass["Sensitivity"]),
    as.numeric(cm_svm$byClass["Sensitivity"])
  ),
  
  Specificity = c(
    as.numeric(cm_lda$byClass["Specificity"]),
    as.numeric(cm_qda$byClass["Specificity"]),
    as.numeric(cm_tree$byClass["Specificity"]),
    as.numeric(cm_rf$byClass["Specificity"]),
    as.numeric(cm_knn$byClass["Specificity"]),
    as.numeric(cm_svm$byClass["Specificity"])
  ),
  
  Balanced_Accuracy = c(
    as.numeric(cm_lda$byClass["Balanced Accuracy"]),
    as.numeric(cm_qda$byClass["Balanced Accuracy"]),
    as.numeric(cm_tree$byClass["Balanced Accuracy"]),
    as.numeric(cm_rf$byClass["Balanced Accuracy"]),
    as.numeric(cm_knn$byClass["Balanced Accuracy"]),
    as.numeric(cm_svm$byClass["Balanced Accuracy"])
  ),
  
  AUC = c(
    as.numeric(auc(roc_lda)),
    as.numeric(auc(roc_qda)),
    as.numeric(auc(roc_tree)),
    as.numeric(auc(roc_rf)),
    as.numeric(auc(roc_knn)),
    as.numeric(auc(roc_svm))
  )
)

results[order(-results$Balanced_Accuracy), ] %>%
  gt() %>%
  fmt_number(
    columns = c(Accuracy, Sensitivity, Specificity, Balanced_Accuracy, AUC),
    decimals = 4
  ) %>%
  tab_header(title = "Model Comparison on Test Set")
```

@tbl-comparision summarises the predictive performance of the six classification models on the test set. While several models achieve relatively high accuracy, this is largely driven by class imbalance and therefore does not fully reflect their ability to identify subscribed clients.

In contrast, metrics such as sensitivity, balanced accuracy, and AUC provide a more informative comparison. The Random Forest model achieves the highest AUC and balanced accuracy, along with the strongest sensitivity, indicating superior overall performance. Decision Tree also demonstrates relatively strong sensitivity, whereas KNN and SVM, despite their high accuracy, show limited ability to detect the minority class.

To further compare the discriminative ability of these models across all classification thresholds, the ROC curves for all six methods are presented in @fig-roc-all-models

```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-cap: "ROC Curve Comparison Across the Six Classification Models"
#| label: fig-roc-all-models
#| fig-align: center
#| fig-width: 7
#| fig-height: 6

library(pROC)
library(ggplot2)
library(dplyr)

roc_to_df <- function(roc_obj, model_name, auc_val) {
  data.frame(
    Specificity = roc_obj$specificities,
    Sensitivity = roc_obj$sensitivities,
    Model = paste0(model_name, " (AUC = ", sprintf("%.4f", auc_val), ")")
  )
}

roc_df <- bind_rows(
  roc_to_df(roc_rf,  "Random Forest", auc(roc_rf)),
  roc_to_df(roc_qda, "QDA",           auc(roc_qda)),
  roc_to_df(roc_knn, "KNN",           auc(roc_knn)),
  roc_to_df(roc_svm, "SVM",           auc(roc_svm)),
  roc_to_df(roc_tree,"Decision Tree", auc(roc_tree)),
  roc_to_df(roc_lda, "LDA",           auc(roc_lda))
)

roc_df$Model <- factor(
  roc_df$Model,
  levels = unique(roc_df$Model)
)

ggplot(roc_df, aes(x = 1 - Specificity, y = Sensitivity, colour = Model)) +
  geom_line(linewidth = 1.3) +
  
  geom_abline(intercept = 0, slope = 1,
              linetype = "dashed",
              colour = "grey50",
              linewidth = 0.8) +
  
  scale_color_manual(values = c(
    "#1b9e77", 
    "#d95f02",  
    "#7570b3", 
    "#e7298a", 
    "#66a61e", 
    "#1f78b4" 
  )) +
  
  labs(
    title = "ROC Curve Comparison Across Models",
    x = "False Positive Rate",
    y = "True Positive Rate",
    colour = NULL
  ) +
  
  theme_minimal(base_size = 13) +
  
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    legend.position = c(0.75, 0.25),
    legend.background = element_rect(fill = "white", colour = "grey80"),
    legend.text = element_text(size = 9),
    legend.spacing.y = unit(0.2, "cm"),
    panel.grid.minor = element_blank()
  )
```

@fig-roc-all-models presents the ROC curves for all six classification models on the test set. The models are ordered according to their AUC values to facilitate comparison. The Random Forest model clearly dominates, with its ROC curve lying above the others for most threshold values, indicating superior discriminative performance. In contrast, although models such as KNN and SVM achieve relatively high accuracy, their ROC curves are consistently lower, reflecting weaker ability to distinguish between subscribed and non-subscribed clients. This visual comparison reinforces the selection of Random Forest as the final model.

## Final Model Evaluation: Random Forest

Based on the comparative analysis, the Random Forest model is selected as the final model due to its superior performance across multiple evaluation metrics. The model is therefore further examined using the test set to assess its predictive capability in detail.

```{r}
#| echo: false
#| warning: false
#| message: false

library(caret)
library(pROC)
pred_rf_opt <- predict(rf_final, newdata = test_data)

prob_rf <- predict(rf_final, newdata = test_data, type = "prob")[, "yes"]

y_test <- test_data$y

cm_rf <- confusionMatrix(pred_rf_opt, y_test, positive = "yes")

roc_rf <- roc(y_test, prob_rf, levels = c("no", "yes"), quiet = TRUE)
```

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-rf-final
#| tbl-cap: "Final Evaluation of the Random Forest Model on the Test Set"

library(gt)

rf_results <- data.frame(
  Metric = c("Accuracy", "Sensitivity", "Specificity", "Balanced Accuracy", "AUC"),
  Value = c(
    as.numeric(cm_rf$overall["Accuracy"]),
    as.numeric(cm_rf$byClass["Sensitivity"]),
    as.numeric(cm_rf$byClass["Specificity"]),
    as.numeric(cm_rf$byClass["Balanced Accuracy"]),
    as.numeric(auc(roc_rf))
  )
)

rf_results %>%
  gt() %>%
  fmt_number(columns = Value, decimals = 4) %>%
  tab_header(title = "Random Forest: Final Test Set Performance")
```

The performance metrics in @tbl-rf-final indicate that the Random Forest model achieves a relatively high overall accuracy (0.8955), primarily driven by its strong ability to correctly classify non-subscribed clients, as reflected in the high specificity (0.9819). However, the sensitivity is comparatively low (0.2237), suggesting that the model still struggles to identify a substantial proportion of subscribed clients.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-rf-cm
#| tbl-cap: "Confusion Matrix for the Final Random Forest Model on the Test Set"

as.data.frame.matrix(cm_rf$table) %>%
  tibble::rownames_to_column("Predicted") %>%
  gt() %>%
  tab_header(title = "Random Forest: Confusion Matrix")
```

This pattern is further confirmed by the confusion matrix, where the number of false negatives (177) is considerably higher than the number of true positives (51). This indicates that many actual subscribers are incorrectly classified as non-subscribers.

Despite this limitation, the model achieves a relatively strong AUC (0.7888), demonstrating good overall discriminative ability. The results suggest that while Random Forest provides the best overall performance among the models considered, there remains room for improvement in detecting the minority class.

# Conclusions {#sec-con}
