---
title: "Project 2: Bank marketing dataset "
author: "Group 17"
number-sections: true
format: 
  html:
    embed-resources: true
editor_options:
  chunk_output_type: console
---

```{r}
#| echo: false
#| warning: false
#| message: false

# load the libraries we will use.

library(skimr)
library(corrplot)
library(gt)
library(dplyr)
library(ggplot2)
library(gridExtra)
```

# Introduction {#sec-intro}

The goal of this project is to predict whether clients will subscribe to a term deposit by analyzing the marketing data of the bank. The research findings of this report can assist financial institutions in formulating more effective marketing strategies, thereby increasing clients conversion rates and reducing marketing costs.

The data used in this study was sourced from a Portuguese bank and covered the detailed information of 17 marketing campaigns conducted by the bank from May 2008 to November 2010. This included the relevant data of 41,118 clients. The dataset included the basic characteristics of the clients, macroeconomic indicators, and variables related to the marketing campaigns. The response variable was a binary variable indicating whether the clients would subscribe to a term deposit ("Yes" or "No"). To enhance the efficiency of the analysis, 10,000 records were randomly selected from the original data as samples for modeling and analysis.

To achieve the research objectives, this study will employ various classification methods, such as KNN, LDA, Tree-based methods, SVM, etc., to model the data and compare the models. These methods are applicable to binary classification problems and can also utilize multiple evaluation metrics to compare the predictive capabilities of the models. Moreover, through the comparison among different models, it is beneficial to better comprehensively understand the relationship between the data features and the prediction results. 

The research results will be compared and analyzed using appropriate evaluation metrics, thereby identifying the model with the best performance and further explaining its predictive capabilities. Ultimately, this will provide a reference basis for future practical applications.

However, this study still has certain limitations. Although there are no missing values in the dataset, the response variable has the problem of class imbalance, which may affect the model's ability to identify minority classes. In the subsequent analysis, appropriate handling of this issue will be considered. Additionally, this data comes from historical marketing activities in a specific period, and the generalization ability of the corresponding analysis results may be limited.

# Exploratory Analysis {#sec-EA}

```{r}
#| echo: false
#| warning: false
#| message: false
#| include: false

## read the data and take a quick look at the data

bank <- read.csv("group_17.csv", header=TRUE)

summary(bank)

#1. Target people for subscription: Primarily young and middle-aged individuals (with an average age of 40), the majority of whom are new  clients (never having been contacted before, most of pdays are 999).
#2. Call situation: There is an obvious variation in call duration (ranging from 1 to 82 (4918/60)minutes), indicating that there are differences in client responses;also, some clients have been contacted excessively (up to 37 times in the campaign).
#3. External economic background: The interest rate fluctuations in this data are relatively obvious (ranging of euribor3m from 0.6% to 5.0%), and sometimes changes in the external economic environment may affect clients' willingness to deposit.

skim(bank)

#1. Data completeness: There are no missing values in these 10,000 pieces of data (complete rate 100%), so the data can be directly used without the need for data filling.
#2. Client diversity: It includes 12 types of occupations and 8 levels of education . The target variable y is a binary classification (yes or no).
#3. Increasing new clients campaign: Over 75% of the clients ( pdays from the 25th percentile to the 100th percentile, all is 999) have never been contacted before, indicating that this is a marketing campaign to increase new clients.
#4. Client situation:
#Main client group: Concentrated in the age range of 32 to 47.
#Call situation: Half of the calls are no longer than 3 minutes (180 seconds).
#Interest rate situation: At the time of data recording, the market was in a high-interest-rate period (the median interest rate was 4.86%).

```

Before making a call, the variable `duration` is usually unknown, and this variable is highly correlated with the target variable. Specifically, it can only be determined whether the client subscribes after the call ends, and longer call duration is usually more likely to result in a subscription. Therefore, adding this variable to the model is equivalent to introducing "future information", which is logically unreasonable and the model has no practical value. Based on these considerations, the `duration` variable will be deleted to ensure that the model remains feasible in practical scenarios.

On the other hand,the dataset is highly imbalanced, with approximately 89% of observations belonging to the "no" class and around 11% to the "yes" class. A random seed (1234) was set, and the dataset was divided into training and test sets using an 80:20 ratio. The class distribution remains consistent after the split, with 88.84% “no” and 11.16% “yes” in the training set, and 88.6% “no” and 11.4% “yes” in the test set. It indicates that the random split has preserved the original class distribution and ensures that both the training and test sets are representative of the overall dataset. This is particularly important in the presence of class imbalance, as it helps avoid biased model evaluation and improves the reliability of performance comparisons across different models.

```{r}
#| echo: false
#| warning: false
#| message: false

#Data preprocessing

bank$duration <- NULL #Removing the "duration" variable is to make the model practical: this variable is unknown before making the call, and it is highly correlated with our target variable (we can know whether the client subscribes after the call, and clients with longer call durations are more likely to subscribe). If it is included in our model, it is equivalent to using "future information", which would lead to logically unreasonable predictions and make the constructed model of no practical value.

#names(bank)

#Change the variable to a 1 or 0 variable indicating whether the client has been contacted or not
bank$pdays_contact <- ifelse(bank$pdays == 999, 0, 1)
bank$pdays <- NULL

# to change the character variables to factors for better fitting
bank$job <- as.factor(bank$job)
bank$marital <- as.factor(bank$marital)
bank$education <- as.factor(bank$education)
bank$default <- as.factor(bank$default)
bank$housing <- as.factor(bank$housing)
bank$loan <- as.factor(bank$loan)
bank$contact <- as.factor(bank$contact)
bank$month <- as.factor(bank$month)
bank$day_of_week <- as.factor(bank$day_of_week)
bank$poutcome <- as.factor(bank$poutcome)
bank$pdays_contact <- as.factor(bank$pdays_contact)
bank$y <- as.factor(bank$y)

```


```{r}
#| echo: false
#| warning: false
#| message: false
#| include: false

# split the data
# set seed
set.seed(1234)

# use 8:2 to split the data to two part (train data and test data)

train_index <- sample(1:nrow(bank), 0.8 * nrow(bank))

train_data <- bank[train_index, ]
test_data  <- bank[-train_index, ]

# check the dimensions
dim(train_data)
dim(test_data)

# check proportion of y
prop.table(table(train_data$y))
prop.table(table(test_data$y))

```


```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-cap: Distribution of Target Variable
#| label: fig-scat1
#| fig-align: center
#| fig-width: 6.5
#| fig-height: 5.5

#table(bank$y)

prop <- prop.table(table(train_data$y)) # to calculate the proportion

bp <- barplot(table(train_data$y),
        main="Distribution of Target Variable", ylim=c(0,8000))
text(bp, table(train_data$y), paste0(round(prop*100, 1), "%"), pos = 3)#to add the proportion to the bar chart for better show in the report 

```

@fig-scat1 illustrates the uneven distribution of the target variable y: the proportion of non-subscribed clients reaches 88.8%, while the proportion of successfully converted clients is only 11.2%. Such class imbalance may introduce bias during model training. Therefore, model performance should not be evaluated solely based on overall accuracy, since even if a model predicts all results as "no", the overall accuracy can still exceed 80%. On the other hand, more attention will be paid to the model's ability to identify these 11.2% key samples, so that the final prediction results have practical reference value.

```{r}
#| echo: false
#| warning: false
#| message: false

# prepare data for correlation analysis (continuous variable)

bank_num <- train_data[,c("age",
                    "campaign",
                    "previous",
                    "emp.var.rate",
                    "cons.price.idx",
                    "cons.conf.idx",
                    "euribor3m",
                    "nr.employed")]

#head(bank_num)


```


```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-cap: Correlation Matrix of Numerical Variables
#| label: fig-scat2
#| fig-align: center
#| fig-width: 6.5
#| fig-height: 5.5

corrplot(cor(bank_num),
         method = "number",#low correlation will be close to white number
         type = "upper", #show half of the plot is enough
         order = "hclust",#to better show the most correlated part
         tl.col = "black")

```


```{r}
#| echo: false
#| warning: false
#| message: false
#| include: false

# The three variables (emp.var.rate, euribor3m, nr.employed) are highly correlated, so we use pca to decide to delete which two variables

bank.pca <- princomp(bank_num, cor = TRUE)

summary(bank.pca)

bank.pca$loadings

```

```{r}
#| echo: false
#| label: tbl-pca
#| tbl-cap: Principal Component Loadings

pca_loadings <- as.data.frame(unclass(bank.pca$loadings))#use unclass() to delete NA

# add variable name
pca_loadings$variable <- rownames(pca_loadings)

# create the gt table
pca_loadings %>%
  select(variable, Comp.1, Comp.2) %>%
  gt() %>%
  fmt_number(
    columns = c(Comp.1, Comp.2),
    decimals = 2 # to keep two decimal places
  ) %>%
  tab_header(
    title = "Principal Component Loadings",
    subtitle = "First two principal components"
  )

```

By combining the correlation matrix @fig-scat2 and Principal Component Analysis @tbl-pca, it can be observed that `euribor3m`, `nr.employed`, and `emp.var.rate` are highly positively correlated (with pairwise correlation coefficients all above 0.9). Multicollinearity is also reflected in the first principal component (Comp.1). To improve subsequent modeling performance, `nr.employed` and `emp.var.rate` are excluded, while `euribor3m` is retained. As the Euro Interbank Offered Rate, `euribor3m` better captures the macro-financial environment during bank marketing and provides stronger interpretability for the final subscription outcomes. Meanwhile, since the variable `default` only has two values, "no" and "unknown", and over 79% of the cases are "no" with no instances of "yes", the available information is extremely limited. Therefore, this variable is removed to reduce noise in the future model.

```{r}
#| echo: false
#| warning: false
#| message: false

# delete nr.employed, and emp.var.rate variable

train_data$nr.employed <- NULL
train_data$emp.var.rate <- NULL
test_data$nr.employed <- NULL
test_data$emp.var.rate <- NULL

#names(train_data)
#names(test_data)
```

```{r}
#| echo: false
#| warning: false
#| message: false
#| include: false

# to have a look at categorical variables
# x and y are all categorical variables, so i use proportion to better show the relationship between them, to see who is more likely to subscribe( set  position = "fill" to show the stacked percentage bar chart)

ggplot(train_data, aes(job, fill = y)) + geom_bar(position = "fill")
# can see clear difference

ggplot(train_data, aes(marital, fill = y)) + geom_bar(position = "fill")
# table(train_data$marital)
# difference is not very clear. Just 13 records are "unknown". So it's possible most of this type is "no".

ggplot(train_data, aes(education, fill = y)) + geom_bar(position = "fill")
# can see clear difference

ggplot(train_data, aes(default, fill = y)) + geom_bar(position = "fill")
# difference is not very clear
# there is no "yes" in the default variable,the information is not enough,so i use table() to decide whether i need to delete it

table(train_data$default)

# Since the variable default only has two values, "no" and "unknown", and over 79% of the cases are "no" with no instances of "yes", the available information is extremely limited. Therefore, we have decided to remove this variable.

train_data$default <- NULL
test_data$default <- NULL

ggplot(train_data, aes(housing, fill = y)) + geom_bar(position = "fill")
# difference is not very clear

ggplot(train_data, aes(loan, fill = y)) + geom_bar(position = "fill")
# difference is not very clear

ggplot(train_data, aes(contact, fill = y)) + geom_bar(position = "fill")
# cellular might be more useful

ggplot(train_data, aes(month, fill = y)) + geom_bar(position = "fill")
# can see clear difference

ggplot(train_data, aes(day_of_week, fill = y)) + geom_bar(position = "fill")
# difference is not very clear

ggplot(train_data, aes(poutcome, fill = y)) + geom_bar(position = "fill")
# can see clear difference

ggplot(train_data, aes(pdays_contact, fill = y)) + geom_bar(position = "fill")
# can see clear difference


```



```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-cap: Subscription Rate Across Key Categorical Variables
#| label: fig-scat3
#| fig-align: center
#| fig-width: 6.5
#| fig-height: 5.5

# choose job, education, poutcome,contact variables to show their relationship with y，use ggplot to do it

p1 <- ggplot(train_data, aes(poutcome, fill = y)) +
  geom_bar(position = "fill") +
  labs(title = "Previous Outcome", x = "poutcome", y = "Proportion") +
  theme_minimal() + #use a simple theme
  theme(axis.text.x = element_text(angle = 45, hjust = 1))#to better show the x label ,rotate it


p2 <- ggplot(train_data, aes(job, fill = y)) +
  geom_bar(position = "fill") +
  labs(title = "Job", x = "job", y = "Proportion") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))#to better show the x label ,rotate it


p3 <- ggplot(train_data, aes(education, fill = y)) +
  geom_bar(position = "fill") +
  labs(title = "Education", x = "education", y = "Proportion") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))#to better show the x label ,rotate it

p4 <- ggplot(train_data, aes(contact, fill = y)) +
  geom_bar(position = "fill") +
  labs(title = "Contact Type", x = "contact", y = "Proportion") +
  theme_minimal()

grid.arrange(p1, p2, p3, p4, ncol = 2,
             top = "Subscription Rate Across Key Categorical Variables")
# use grid.arrange to put four graphs in one plot

```

@fig-scat3 shows the relationships between several key categorical variables and the variable indicating whether the client subscribes (y). Due to the large number of categorical variables, only selected variables with stronger interpretability and potential relevance to the target variable are presented. From the graphs, it can be seen that the previous marketing outcome (`poutcome`) obviously affects whether the client subscribes this time. Clients who were successfully marketed before (success) have a higher probability of subscribing than those who were marketing failed (failure) or had no previous contact (nonexistent). Moreover, there are certain differences between different occupations (`job`) and educational levels (`education`), indicating that the social characteristics of the clients may have an impact on the subscription decision. At the same time, there are also certain differences in contact methods (`contact`), with clients contacted via mobile phone (cellular) having a slightly higher chance of subscribing than those contacted via telephone. 

Overall, historical campaign outcomes, client social characteristics, and communication channels may all have an impact on whether the client subscribes this time. Therefore, these variables will have certain predictive value in the subsequent modeling process.

```{r}
#| echo: false
#| warning: false
#| message: false
#| include: false

# to see the selected variables
names(train_data)
names(test_data)
```

```{r}
#| include: false
#| echo: false
#| label: tbl-var
#| tbl-cap: Variables Used in the Analysis

# This table presents the variables that we ultimately used for the analysis.

# use the information in Table 2: Variable description of bank marketing dataset to create a dataframe for better showing in the report .But it's too much. So we decided to not show it in the report, instead we use "Feature Selection Summary" to explain this part.

var_table <- data.frame(
  Variable = c(
    "y",
    "age","job","marital","education","housing","loan",
    "contact","month","day_of_week",
    "campaign","pdays_contact","previous","poutcome",
    "cons.price.idx","cons.conf.idx","euribor3m"
  ),
  
  Type = c(
    "Categorical",
    "Numeric","Categorical","Categorical","Categorical","Categorical","Categorical",
    "Categorical","Categorical","Categorical",
    "Numeric","Categorical","Numeric","Categorical",
    "Numeric","Numeric","Numeric"
  ),
  
  Description = c(
    "Has the client subscribed a term deposit (yes/no)",
    
    "Age of the client at the contact date",
    "Type of job (e.g., admin, technician, services, etc.)",
    "Marital status (divorced, married, single)",
    "Education level (basic.4y, basic.6y, basic.9y, high.school, university.degree, etc.)",
    "Whether the client has a housing loan",
    "Whether the client has a personal loan",
    
    "Contact communication type (cellular or telephone)",
    "Last contact month of the year",
    "Last contact day of the week",
    
    "Number of contacts performed during this campaign for this client",
    "Indicator whether the client has been contacted previously",
    "Number of contacts performed before this campaign",
    "Outcome of the previous marketing campaign (failure, nonexistent, success)",
    
    "Consumer price index (monthly indicator)",
    "Consumer confidence index (monthly indicator)",
    "Euribor 3-month rate (daily indicator)"
  )
)

# use gt() to better show the selected variable
var_table %>%
  gt() %>%
  tab_header(
    title = "Variables Used in the Analysis",
    subtitle = "Selected variables from the Bank Marketing dataset"
  )  %>% # add title and subtitle
  cols_label(
    Variable = "Variable Name",
    Type = "Variable Type",
    Description = "Description"
  ) #change the column label
```
### Feature Selection Summary {#sec-FSS .unnumbered}

The final set of variables includes clients characteristics (e.g., `age`, `job`, `education`), campaign-related variables (e.g., `campaign`, `previous`, `poutcome`), and macroeconomic indicators (e.g., `euribor3m`, consumer indices). Variables with no practical value or low information content, such as duration and default, will be excluded. Additionally, multiple collinearity issues are addressed by retaining representative variables. Overall, the selected features lay a solid foundation for the subsequent modeling work.

# Formal Analysis {#sec-FA}

# Conclusions {#sec-con}


