---
title: "KNN_SVM"
format: html
editor: visual
---
```{r}
#| echo: false
#| warning: false
#| message: false

# load the libraries we will use.

library(skimr)
library(corrplot)
library(gt)
library(dplyr)
library(ggplot2)
library(gridExtra)
```

# Introduction {#sec-intro}

# Exploratory Analysis {#sec-EA}

```{r}
#| echo: false
#| warning: false
#| message: false

## read the data and take a quick look at the data

bank <- read.csv("group_17.csv", header=TRUE)

summary(bank)

#1. Target people for subscription: Primarily young and middle-aged individuals (with an average age of 40), the majority of whom are new  clients (never having been contacted before).
#2. Call situation: There is an obvious variation in call duration (ranging from 1 to 82 minutes), indicating that there are differences in client responses;also, some clients have been contacted excessively (up to 37 times).
#3. External economic background: The interest rate fluctuations in this data are relatively obvious (ranging from 0.6% to 5.0%), and sometimes changes in the external economic environment may affect clients' willingness to deposit.

skim(bank)

#1. Data completeness: There are no missing values in these 10,000 pieces of data (complete rate 100%), so the data can be directly used without the need for data filling.
#2. Client diversity: It includes 12 types of occupations and 8 levels of education . The target variable y is a binary classification (yes or no).
#3. Increasing new clients campaign: Over 75% of the clients ( pdays from the 25th percentile to the 100th percentile, all is 999) have never been contacted before, indicating that this is a marketing campaign to increase new clients.
#4. Client situation:
#Main client group: Concentrated in the age range of 32 to 47.
#Call situation: Half of the calls are no longer than 3 minutes (180 seconds).
#Interest rate situation: At the time of data recording, the market was in a high-interest-rate period (the median interest rate was 4.86%).

```

```{r}
#| echo: false
#| warning: false
#| message: false

#Data preprocessing

bank$duration <- NULL #Removing the "duration" variable is to make the model practical: this variable is unknown before making the call, and it is highly correlated with our target variable (we can know whether the client subscribes after the call, and clients with longer call durations are more likely to subscribe). If it is included in our model, it is equivalent to using "future information", which would lead to logically unreasonable predictions and make the constructed model of no practical value.

#names(bank)

#Change the variable to a 1 or 0 variable indicating whether the client has been contacted or not
bank$pdays_contact <- ifelse(bank$pdays == 999, 0, 1)
bank$pdays <- NULL

# to change the character variables to factors for better fitting
bank$job <- as.factor(bank$job)
bank$marital <- as.factor(bank$marital)
bank$education <- as.factor(bank$education)
bank$default <- as.factor(bank$default)
bank$housing <- as.factor(bank$housing)
bank$loan <- as.factor(bank$loan)
bank$contact <- as.factor(bank$contact)
bank$month <- as.factor(bank$month)
bank$day_of_week <- as.factor(bank$day_of_week)
bank$poutcome <- as.factor(bank$poutcome)
bank$pdays_contact <- as.factor(bank$pdays_contact)
bank$y <- as.factor(bank$y)

```

@fig-scat1 This chart illustrates the uneven distribution of the target variable y: the proportion of non-subscribed clients reaches 88.8%, while the proportion of successfully converted clients is only 11.2%. Due to this distribution characteristic, our subsequent logistic regression model may exhibit certain bias during the fitting process. Therefore, when evaluating the model's performance, we do not solely rely on the overall accuracy rate (because even if the model predicts all results as "no", the final accuracy rate can still exceed 80%). On the other hand, we will pay more attention to the model's ability to identify these 11.21% key samples, so that the final prediction results have practical reference value.

```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-cap: Distribution of Target Variable
#| label: fig-scat1
#| fig-align: center
#| fig-width: 6.5
#| fig-height: 5.5

#table(bank$y)

prop <- prop.table(table(bank$y))

bp <- barplot(table(bank$y),
        main="Distribution of Target Variable", ylim=c(0,10000))
text(bp, table(bank$y), paste0(round(prop*100, 1), "%"), pos = 3)#to add the proportion to the bar chart for better show in the report 

```


```{r}
#| echo: false
#| warning: false
#| message: false

# prepare data for correlation analysis (continuous variable)

bank_num <- bank[,c("age",
                    "campaign",
                    "previous",
                    "emp.var.rate",
                    "cons.price.idx",
                    "cons.conf.idx",
                    "euribor3m",
                    "nr.employed")]

#head(bank_num)


```


```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-cap: Correlation Matrix of Numerical Variables
#| label: fig-scat2
#| fig-align: center
#| fig-width: 6.5
#| fig-height: 5.5

corrplot(cor(bank_num),
         method = "number",#low correlation will be close to white number
         type = "upper", #show half of the plot is enough
         order = "hclust",#to better show the most correlated part
         tl.col = "black")

```


```{r}
#| echo: false
#| warning: false
#| message: false

# The three variables (emp.var.rate, euribor3m, nr.employed) are highly correlated, so we use pca to decide to delete which two variables

bank.pca <- princomp(bank_num, cor = TRUE)

#summary(bank.pca)

#bank.pca$loadings

```

```{r}
#| echo: false
#| label: tbl-pca
#| tbl-cap: Principal Component Loadings

pca_loadings <- as.data.frame(unclass(bank.pca$loadings))#use unclass() to delete NA

# add variable name
pca_loadings$variable <- rownames(pca_loadings)

# create the gt table
pca_loadings %>%
  select(variable, Comp.1, Comp.2) %>%
  gt() %>%
  fmt_number(
    columns = c(Comp.1, Comp.2),
    decimals = 2 # to keep two decimal places
  ) %>%
  tab_header(
    title = "Principal Component Loadings",
    subtitle = "First two principal components"
  )

```

By combining the correlation matrix @fig-scat2 and Principal Component Analysis @tbl-pca, it can be observed that euribor3m, nr.employed, and emp.var.rate have a high degree of positive correlation (the pairwise correlation coefficients are all greater than 0.9), and multiple collinearity can also be seen in the first principal component (Comp.1). For the subsequent modeling performance, we decided to exclude nr.employed and emp.var.rate and retain euribor3m. Because euribor3m represents the Euro Interbank Offered Rate, it can better reflect the macro-financial situation during bank marketing, and can relatively better help us explain the final subscription situation of clients.

```{r}
#| echo: false
#| warning: false
#| message: false

# delete nr.employed, and emp.var.rate variable

bank$nr.employed <- NULL
bank$emp.var.rate <- NULL

#names(bank)

```

```{r}
#| echo: false
#| warning: false
#| message: false

# to have a look at category variable
# x and y are all category variable ,so i use proportion to better show the relationship between them,to see who is more likely to subscribe( set  position = "fill" to show the proportion barplot)

ggplot(bank, aes(job, fill = y)) + geom_bar(position = "fill")
# can see clear difference

ggplot(bank, aes(marital, fill = y)) + geom_bar(position = "fill")
# difference is not very clear

ggplot(bank, aes(education, fill = y)) + geom_bar(position = "fill")
# can see clear difference

ggplot(bank, aes(default, fill = y)) + geom_bar(position = "fill")
# difference is not very clear
# there is no "yes" in the default variable,the information is not enough,so i use table() to decide whether i need to delete it

#table(bank$default)

# Since the variable default only has two values, "no" and "unknown", and over 78% of the cases are "no" with no "yes", the available information is extremely limited. Therefore, we have decided to remove this variable.

bank$default <- NULL

ggplot(bank, aes(housing, fill = y)) + geom_bar(position = "fill")
# difference is not very clear

ggplot(bank, aes(loan, fill = y)) + geom_bar(position = "fill")
# difference is not very clear

ggplot(bank, aes(contact, fill = y)) + geom_bar(position = "fill")
# cellular might be more useful

ggplot(bank, aes(month, fill = y)) + geom_bar(position = "fill")
# can see clear difference

ggplot(bank, aes(day_of_week, fill = y)) + geom_bar(position = "fill")
# difference is not very clear

ggplot(bank, aes(poutcome, fill = y)) + geom_bar(position = "fill")
# can see clear difference

ggplot(bank, aes(pdays_contact, fill = y)) + geom_bar(position = "fill")
# can see clear difference


```

Since the variable default only has two values, "no" and "unknown", and over 78% of the cases are "no" with no "yes", the available information is extremely limited. Therefore, we have decided to remove this variable.

```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-cap: Subscription Rate Across Key Categorical Variables
#| label: fig-scat3
#| fig-align: center
#| fig-width: 6.5
#| fig-height: 5.5

# choose job, education, poutcome,contact variables to show their relationship with y，use ggplot to do it

p1 <- ggplot(bank, aes(poutcome, fill = y)) +
  geom_bar(position = "fill") +
  labs(title = "Previous Outcome", x = "poutcome", y = "Proportion") +
  theme_minimal() + #use a simple theme
  theme(axis.text.x = element_text(angle = 45, hjust = 1))#to better show the x label ,rotate it


p2 <- ggplot(bank, aes(job, fill = y)) +
  geom_bar(position = "fill") +
  labs(title = "Job", x = "job", y = "Proportion") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))#to better show the x label ,rotate it


p3 <- ggplot(bank, aes(education, fill = y)) +
  geom_bar(position = "fill") +
  labs(title = "Education", x = "education", y = "Proportion") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))#to better show the x label ,rotate it

p4 <- ggplot(bank, aes(contact, fill = y)) +
  geom_bar(position = "fill") +
  labs(title = "Contact Type", x = "contact", y = "Proportion") +
  theme_minimal()

grid.arrange(p1, p2, p3, p4, ncol = 2,
             top = "Subscription Rate Across Key Categorical Variables")
# use grid.arrange to put four graphs in one plot

```

@fig-scat3 This set of graphs respectively shows the relationships between several key categorical variables and the variable indicating whether the client subscribes (y). From the graphs, it can be seen that the previous marketing outcome (poutcome) obviously affects whether the client subscribes this time. Clients who were successfully marketed before (success) have a higher probability of subscribing than those who were marketing failed (failure) or had no previous contact (nonexistent). Moreover, there are certain differences between different occupations (job) and educational levels (education), indicating that the social characteristics of the clients may have an impact on the subscription decision. At the same time, there are also certain differences in contact methods (contact), with clients contacted via mobile phone (cellular) having a slightly higher chance of subscribing than those contacted via telephone (telephone). 

Overall, historical marketing results, client social characteristics, and communication channels may all have an impact on whether the client subscribes this time. Therefore, these variables will have certain predictive value in the subsequent modeling process.

```{r}
#| echo: false
#| warning: false
#| message: false

# to see the selected variables
names(bank)

```

```{r}
#| echo: false
#| label: tbl-var
#| tbl-cap: Variables Used in the Analysis

# use the information in Table 2: Variable description of bank marketing dataset to create a dataframe for better showing in the report
var_table <- data.frame(
  Variable = c(
    "y",
    "age","job","marital","education","housing","loan",
    "contact","month","day_of_week",
    "campaign","pdays_contact","previous","poutcome",
    "cons.price.idx","cons.conf.idx","euribor3m"
  ),
  
  Type = c(
    "Categorical",
    "Numeric","Categorical","Categorical","Categorical","Categorical","Categorical",
    "Categorical","Categorical","Categorical",
    "Numeric","Categorical","Numeric","Categorical",
    "Numeric","Numeric","Numeric"
  ),
  
  Description = c(
    "Has the client subscribed a term deposit (yes/no)",
    
    "Age of the client at the contact date",
    "Type of job (e.g., admin, technician, services, etc.)",
    "Marital status (divorced, married, single)",
    "Education level (basic.4y, basic.6y, basic.9y, high.school, university.degree, etc.)",
    "Whether the client has a housing loan",
    "Whether the client has a personal loan",
    
    "Contact communication type (cellular or telephone)",
    "Last contact month of the year",
    "Last contact day of the week",
    
    "Number of contacts performed during this campaign for this client",
    "Indicator whether the client has been contacted previously",
    "Number of contacts performed before this campaign",
    "Outcome of the previous marketing campaign (failure, nonexistent, success)",
    
    "Consumer price index (monthly indicator)",
    "Consumer confidence index (monthly indicator)",
    "Euribor 3-month rate (daily indicator)"
  )
)

# use gt() to better show the selected variable
var_table %>%
  gt() %>%
  tab_header(
    title = "Variables Used in the Analysis",
    subtitle = "Selected variables from the Bank Marketing dataset"
  )  %>% # add title and subtitle
  cols_label(
    Variable = "Variable Name",
    Type = "Variable Type",
    Description = "Description"
  ) #change the column label
```

@tbl-var This table presents the variables that we ultimately used for the analysis.

```{r}
#| echo: false
#| warning: false
#| message: false

# split the data
# set seed
set.seed(1234)

# use 8:2 to split the data to two part (train data and test data)

train_index <- sample(1:nrow(bank), 0.8 * nrow(bank))

train_data <- bank[train_index, ]
test_data  <- bank[-train_index, ]

# check the dimensions
dim(train_data)
dim(test_data)

# check proportion of y
prop.table(table(train_data$y))
prop.table(table(test_data$y))

```


```{r}
#| echo: false
#| warning: false
#| message: false

packages <- c(
  "caret",
  "pROC",
  "gt",
  "ggplot2",
  "gridExtra",
  "class",
  "kernlab"
)

for(pkg in packages){
  library(pkg, character.only = TRUE)
}
```

```{r}
#| echo: false
#| warning: false
#| message: false

# Shared preprocessing for KNN and SVM
# These two methods require numerical inputs, so factor variables are converted
# into dummy variables and then standardised using training-set statistics only.

x_train_raw <- subset(train_data, select = -y)
x_test_raw  <- subset(test_data,  select = -y)

# create dummy variables from factor predictors
dummy_obj <- dummyVars(~ ., data = x_train_raw)

x_train_mat <- predict(dummy_obj, newdata = x_train_raw)
x_test_mat  <- predict(dummy_obj, newdata = x_test_raw)

# standardise using training set mean and sd
preproc_obj <- preProcess(x_train_mat, method = c("center", "scale"))

x_train_scaled <- predict(preproc_obj, x_train_mat)
x_test_scaled  <- predict(preproc_obj, x_test_mat)

# outcome variables
y_train <- train_data$y
y_test  <- test_data$y

```

# KNN
## Model description and justification
The k-Nearest Neighbours (KNN) classifier is a non-parametric method that assigns class labels based on the majority class among the k closest observations in the predictor space. As KNN relies on distance calculations, all predictors were converted into numerical form through dummy encoding and subsequently standardised to ensure comparability across variables.

KNN was included in this study as a flexible, data-driven benchmark that does not impose a parametric structure on the relationship between predictors and the response. This makes it particularly useful for capturing local patterns in the data and for comparison with more structured classification methods.

```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-cap: "Cross-Validated ROC for KNN Across Candidate k Values"
#| label: fig-knn-tune
#| fig-align: center
#| fig-width: 6
#| fig-height: 4.5

set.seed(1234)

# caret's twoClassSummary assumes the first level is the event of interest,
# so reorder levels to make "yes" the positive class during tuning
y_train_knn <- factor(y_train, levels = c("yes", "no"))

ctrl_knn <- trainControl(
  method = "cv",
  number = 5,
  classProbs = TRUE,
  summaryFunction = twoClassSummary
)

knn_grid <- expand.grid(k = c(3, 5, 7, 9, 11, 15, 21))

knn_fit <- train(
  x = x_train_scaled,
  y = y_train_knn,
  method = "knn",
  metric = "ROC",
  tuneGrid = knn_grid,
  trControl = ctrl_knn
)

ggplot(knn_fit$results, aes(x = k, y = ROC)) +
  geom_line() +
  geom_point(size = 2) +
  theme_minimal() +
  labs(x = "Number of Neighbours (k)",
       y = "Cross-Validated ROC",
       title = "KNN Tuning")
```

## Hyperparameter tuning
The primary tuning parameter for KNN is the number of neighbours, k. To determine an appropriate value, a range of candidate values was evaluated using cross-validated ROC as the selection criterion. This approach ensures that the chosen model maximises its ability to discriminate between subscribed and non-subscribed clients.

The results indicate that model performance improves as k increases and stabilises beyond moderate values. The highest cross-validated ROC was achieved at k = 21, which was therefore selected as the optimal hyperparameter. Although performance differences beyond k = 9 were relatively small, selecting k = 21 ensures that the model achieves its best observed predictive performance.
```{r}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-knn-best
#| tbl-cap: "KNN: Selected Hyperparameter"

knn_best <- data.frame(
  Hyperparameter = "Best k",
  Value = knn_fit$bestTune$k
)

knn_best %>%
  gt() %>%
  tab_header(title = "KNN: Selected Hyperparameter")
```

## Model Evaluation
The selected KNN model was evaluated on the test set using Accuracy, Sensitivity, Specificity, Balanced Accuracy, and the Area Under the ROC Curve (AUC). While the model achieved a high overall accuracy, this result is influenced by the strong class imbalance in the dataset.

More informative metrics, such as Sensitivity and AUC, indicate that the model has limited ability to correctly identify subscribed clients, which constitute the minority class. This suggests that, although KNN performs reasonably well in overall classification, its effectiveness in identifying the key target group is restricted.
```{r}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-knn-results
#| tbl-cap: "KNN: Performance on Test Set"

# predictions on the test set
pred_knn <- predict(knn_fit, x_test_scaled)
prob_knn <- predict(knn_fit, x_test_scaled, type = "prob")[, "yes"]

# reorder prediction factor to match original y levels for confusion matrix
pred_knn <- factor(pred_knn, levels = c("no", "yes"))

cm_knn  <- confusionMatrix(pred_knn, y_test, positive = "yes")
roc_knn <- roc(y_test, prob_knn, levels = c("no", "yes"), quiet = TRUE)

knn_results <- data.frame(
  Metric = c("Accuracy", "Sensitivity", "Specificity",
             "Balanced Accuracy", "AUC"),
  Value = c(
    round(cm_knn$overall["Accuracy"], 4),
    round(cm_knn$byClass["Sensitivity"], 4),
    round(cm_knn$byClass["Specificity"], 4),
    round(cm_knn$byClass["Balanced Accuracy"], 4),
    round(auc(roc_knn), 4)
  )
)

knn_results %>%
  gt() %>%
  cols_label(Metric = "Metric", Value = "Value") %>%
  tab_header(title = "KNN: Performance on Test Set")
```

# SVM
## Model description and justification
Support Vector Machines (SVM) classify observations by identifying an optimal separating boundary that maximises the margin between classes. In this study, a radial basis function (RBF) kernel was employed to allow for nonlinear decision boundaries, enabling the model to capture complex relationships between predictors and the response variable.

SVM was selected due to its strong performance in high-dimensional settings and its ability to construct stable classification boundaries after dummy encoding of categorical variables. Compared with KNN, SVM provides a more structured approach to classification and can offer improved generalisation in the presence of noisy predictors.

```{r}
#| echo: false
#| warning: false
#| message: false

set.seed(1234)

y_train_svm <- factor(y_train, levels = c("yes", "no"))

ctrl_svm <- trainControl(
  method = "cv",
  number = 3,
  classProbs = TRUE,
  summaryFunction = twoClassSummary
)

svm_grid <- expand.grid(
  sigma = c(0.01, 0.05),
  C     = c(0.5, 1, 2)
)

svm_fit <- train(
  x = x_train_scaled,
  y = y_train_svm,
  method = "svmRadial",
  metric = "ROC",
  tuneGrid = svm_grid,
  trControl = ctrl_svm
)

svm_fit$bestTune
svm_fit$results
```

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-SVM-turning
ggplot(svm_fit$results, aes(x = factor(C), y = factor(sigma), fill = ROC)) +
  geom_tile(color = "white") +
  theme_minimal() +
  labs(x = "Cost (C)",
       y = "Sigma",
       fill = "ROC",
       title = "SVM Tuning")
```
@fig-SVM-turning illustrates the cross-validated ROC values for different combinations of the SVM hyperparameters, namely the cost parameter (C) and the kernel parameter (sigma). Each cell represents the predictive performance of a specific parameter combination, with darker shading indicating higher ROC values.

It can be observed that models with a larger sigma (0.05) consistently achieve higher ROC values compared to those with sigma = 0.01, suggesting that a more flexible decision boundary improves classification performance in this dataset. Among all combinations, the highest ROC is obtained at sigma = 0.05 and C = 0.5, indicating that this parameter setting provides the best trade-off between model complexity and generalisation ability.
```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-SVM-turning-results
ggplot(svm_fit$results, aes(x = factor(C), y = ROC, colour = factor(sigma), group = factor(sigma))) +
  geom_line() +
  geom_point(size = 2) +
  theme_minimal() +
  labs(
    x = "Cost (C)",
    y = "Cross-Validated ROC",
    colour = "Sigma",
    title = "SVM Tuning Results"
  )
```
@fig-SVM-turning-results presents the variation in cross-validated ROC across different values of the cost parameter (C) for two levels of sigma. Each line corresponds to a fixed sigma value, allowing a direct comparison of how model performance changes with C under different levels of model flexibility.

For both sigma values, the ROC slightly decreases as C increases, indicating that overly large cost values may lead to overfitting and reduced generalisation performance. Furthermore, the model with sigma = 0.05 consistently outperforms that with sigma = 0.01 across all values of C. The highest ROC is achieved at sigma = 0.05 and C = 0.5, confirming the result observed in the heatmap and supporting the selection of this parameter combination as the optimal SVM model.

## Hyperparameter tuning
```{r}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-svm-best
#| tbl-cap: "SVM: Selected Hyperparameters"

svm_best <- data.frame(
  Hyperparameter = c("Best sigma", "Best cost (C)"),
  Value = c(svm_fit$bestTune$sigma, svm_fit$bestTune$C)
)

svm_best %>%
  gt() %>%
  tab_header(title = "SVM: Selected Hyperparameters")
```
The performance of the tuned SVM model was assessed on the test set using the same evaluation metrics as for KNN. The model achieved high overall accuracy, but, similar to KNN, this result is largely driven by the dominance of the majority class.

The relatively low sensitivity indicates that the SVM model struggles to correctly identify subscribed clients, suggesting that the decision boundary may favour the majority class. The AUC value provides a more balanced assessment of performance and indicates that the SVM model achieves moderate discrimination ability, though slightly lower than that of KNN in this application.


## Model Evaluation
```{r}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-svm-results
#| tbl-cap: "SVM: Performance on Test Set"

pred_svm <- predict(svm_fit, x_test_scaled)
prob_svm <- predict(svm_fit, x_test_scaled, type = "prob")[, "yes"]

pred_svm <- factor(pred_svm, levels = c("no", "yes"))

cm_svm  <- confusionMatrix(pred_svm, y_test, positive = "yes")
roc_svm <- roc(y_test, prob_svm, levels = c("no", "yes"), quiet = TRUE)

svm_results <- data.frame(
  Metric = c("Accuracy", "Sensitivity", "Specificity",
             "Balanced Accuracy", "AUC"),
  Value = c(
    round(cm_svm$overall["Accuracy"], 4),
    round(cm_svm$byClass["Sensitivity"], 4),
    round(cm_svm$byClass["Specificity"], 4),
    round(cm_svm$byClass["Balanced Accuracy"], 4),
    round(auc(roc_svm), 4)
  )
)

svm_results %>%
  gt() %>%
  cols_label(Metric = "Metric", Value = "Value") %>%
  tab_header(title = "SVM: Performance on Test Set")
```

